---
author:
  Sean Davis, MD/PhD
title: "The Hou*R* of Code"
subtitle: 'An overview of the *R* language in action'
date: '`r format(Sys.Date(),"%A, %B %d, %Y")`'
output: 
  ioslides_presentation:
    keep_md: true
---

# Introduction

## What is R?

- A software package
- A programming language
- A toolkit for developing statistical and analytical tools
- An extensive library of statistical and mathematical software and algorithms
- A scripting language
- ...

## Why R?

- R is cross-platform and runs on Windows, Mac, and Linux (as well as more obscure systems).
- R provides a vast number of useful statistical tools, many of which have been painstakingly tested.
- R produces publication-quality graphics in a variety of formats.
- R plays well with FORTRAN, C, and scripts in many languages.
- R scales, making it useful for small and large projects.  It is NOT Excel.
- R eschews the GUI.

> I can develop code for analysis on my Mac laptop.  I can then install the *same* code on our massive computer cluster and run it in parallel on 1000 samples, monitor the process, and then update a database with R when complete.

## Why not R?

- R cannot do everything.
- R is not always the ``best'' tool for the job.
- R will \textit{not hold your hand.
- The documentation can be opaque.
- R can drive you crazy (on a good day) or age you prematurely (on a bad one).
- Finding the right package to do the job you want to do can be challenging; worse, some contributed packages are unreliable.
- R eschews the GUI.

## R License and the Open Source Ideal

- R is free!
- Distributed under GNU license
    - You may download the source code.
    - You may modify the source code to your heart's content.
    - You may distribute the modified source code and even charge money for it, 
    - *but* you must distribute the modified source code under the original GNU license

### Take-home Message

This license means that R will always be available, will always be open source, and can grow organically without constraint.

# Getting started

## Installation

The first step is to install R. You can download and install R from
the [Comprehensive R Archive Network](https://cran.r-project.org/)
(CRAN). It is relatively straightforward, but if you need further help
you can try the following resources: 

* [Installing R on Windows](https://github.com/genomicsclass/windows#installing-r) 
* [Installing R on Mac](http://youtu.be/Icawuhf0Yqo)
* [Installing R on Ubuntu](http://cran.r-project.org/bin/linux/ubuntu/README)

## Installing RStudio

The next step is to install [RStudio](https://www.rstudio.com), a program for viewing and running
R scripts. Technically you can run all the code shown here without
installing RStudio, but we highly recommend this integrated
development environment (IDE). 

- [Instructions are here](http://www.rstudio.com/products/rstudio/download/). 

## R on your own

Try out the [swirl](http://swirlstats.com/) tutorial, which teaches you R programming and data science interactively, at your own pace and in the R console. Once you have R installed, you can install `swirl` and run it the following way: 

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

## Quick R references

There are also many open and free resources and reference
guides for R. Two examples are:

- [Quick-R](http://www.statmethods.net/): a quick online reference for data input, basic statistics and plots
- R reference card [PDF](https://cran.r-project.org/doc/contrib/Short-refcard.pdf) by Tom Short

# R from zero

## Follow along

- In RStudio, copy and paste the following:

```{r eval=FALSE}
download.file('https://raw.githubusercontent.com/seandavi/hour_of_code/master/HourOfCode.Rmd',
              destfile='HourOfCode.Rmd')
```

- In the file pane, you can choose the `HourOfCode.Rmd` file and it will open in the RStudio text pane. 


## Interacting with R

### Expression

```{r}
1 + pi + sin(3.7)
```

### Assignment

```{r}
x = 1
y <- 2
3 -> z
```

## Interacting with R

- The `<-`, `->` and `=` are all assignment operators.

```{r eval=FALSE}
x = 1
y <- 2
3 -> z
```

- If a line is not a complete R command, R will continue the next line with a `+`.

```{r eval=FALSE}
1 + pi + 
    sin(3.7)
```

## Getting help

R has extensive help functionality built in.

```{r eval=FALSE}
help('print')
help(print)
?print
?data.frame
?`+`
```

```{r eval=FALSE}
help.search('microarray')
RSiteSearch('microarray')
```

- *For any new function that you see, type `help(newfunction)`*.

# First steps in R

## Paths and the Working Directory

When you are working in R it is useful to know your _working directory_. This is the directory or folder in which R will save or look for files by default. You can see your working directory by typing:

```{r, eval=FALSE}
getwd()
```

## Loading data into R

R can read files of many different types and from many different sources. 

### Directly from the web

```{r}
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
url <- paste0(dir, "femaleMiceWeights.csv")
dat <- read.csv(url)
```

### Download first

```{r eval=FALSE}
library(downloader) ##use install.packages to install
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
filename <- "femaleMiceWeights.csv" 
url <- paste0(dir, filename)
if (!file.exists(filename)) download(url, destfile=filename)
```

## Working with data

```{r eval=FALSE}
head(dat)
tail(dat)
summary(dat)
dim(dat)
```

## Working with data

```{r}
head(dat)
```

## Working with data

```{r}
tail(dat)
```

## Working with data

```{r}
summary(dat)
```

## Working with data

```{r}
dim(dat)
```

# dplyr

## dplyr filter

```{r}
library(dplyr) 
chow <- filter(dat, Diet=="chow") #keep only the ones with chow diet
head(chow)
```

## dplyr select

```{r}
chowVals <- select(chow,Bodyweight)
head(chowVals)
```

## Piping

```{r}
chowVals <- filter(dat, Diet=="chow") %>% select(Bodyweight) %>% unlist
hist(chowVals)
```

# Plotting with ggplot2

## ggplot2 package

The `ggplot2` package is a relatively novel approach to generating highly informative publication-quality graphics. The "gg" stands for "Grammar of Graphics". In short, instead of thinking about a single function that produces a plot, `ggplot2` uses a "grammar" approach, akin to building more and more complex sentences to layer on more information or nuance. 

## Data Model

The `ggplot2` package assumes that data are in the form of a data.frame. In some cases, the data will need to be manipulated into a form that matches assumptions that `ggplot2` uses.  In particular, if one has a *matrix* of numbers associated with different subjects (samples, people, etc.), the data will usually need to be transformed into a "long" data frame. 

## Getting started

To use the `ggplot2` package, it must be installed and loaded. Assuming that installation has been done already, we can load the package directly:

```{r load}
library(ggplot2)
```

## Playing with ggplot2

## mtcars data

We are going to use the mtcars dataset, included with R, to experiment with `ggplot2`. 

```{r mtcarsLoad}
data(mtcars)
```

- Exercise: Explore the `mtcars` dataset using `View`, `summary`, `dim`, `class`, etc.

## Pairs plot

We can also take a quick look at the relationships between the variables using the `pairs` plotting function.

```{r mtcarsPlot}
pairs(mtcars[,1:4])
```

## Go to the vignette

```{r eval=FALSE}
download.file('https://raw.githubusercontent.com/seandavi/hour_of_code/master/ggplot2.Rmd',
              destfile='ggplot2.Rmd')
```

# Reproducibility

## Literate programming

- See the [code for these slides](https://github.com/seandavi/hour_of_code/blob/master/HourOfCode.Rmd)
- [R markdown](https://rmarkdown.rstudio.com/)

## sessionInfo

```{r}
sessionInfo()
```
